---
title: "HW 03"
author: "Sravani Maddineni"
format:
  html:
    embed-resources: true
execute :
  warning: false
  message: false
  error: false
toc: true
---

## 1 - Du Bois challenge.

```{r}
# label: label-me-1
# Load libraries
library(ggplot2)
library(tidyr)
library(readr)
library(dplyr)
library(stringr)
library(tinytex)
library(jpeg)
library(grid)
library(scales)



library(cowplot)

# Load the background image
bg_img <- readJPEG("C:/Users/Reliance/Desktop/dataviz/hw-03/parchment.jpeg")

# Load and prepare the data
data <- read_csv("C:/Users/Reliance/Desktop/dataviz/hw-03/income.csv")
income_long <- data %>%
  pivot_longer(cols = Rent:Other, names_to = "Category", values_to = "Expenditure") %>%
  mutate(Category = factor(Category, levels = c("Rent", "Food", "Clothes", "Tax", "Other"), ordered = TRUE)) %>%
  mutate(Class_Label = paste("CLASS: ", Class, "\nACTUAL AVERAGE: ", Average_Income))

# Calculate total expenditure for each class
total_expenditure <- income_long %>%
  group_by(Class_Label) %>%
  summarise(Total = sum(Expenditure), .groups = 'drop')

# Join total expenditure back to income_long
income_long <- income_long %>%
  left_join(total_expenditure, by = "Class_Label") %>%
  mutate(Percentage = (Expenditure / Total) * 100)

# Define category colors
category_colors <- c(
  Rent = "black",
  Food = "purple",
  Clothes = "pink",
  Tax = "lightblue",
  Other = "grey"
)

# Create the main plot
p <- ggplot(income_long, aes(x = Expenditure, y = reorder(Class_Label, -Average_Income), fill = Category)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE), width = 0.7) + 
  geom_text(aes(label = ifelse(Percentage >= 1.5, paste0(round(Percentage, 1), "%"), "")), 
            position = position_stack(reverse = TRUE, vjust = 0.5), size = 2.5, 
            color = ifelse(income_long$Category == "Rent", "white", "black")) +
  scale_fill_manual(values = category_colors) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "transparent", color = NA),
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    axis.title.y = element_blank(),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    legend.title = element_blank(),
    axis.title.x = element_blank(),
    legend.text = element_blank(),
    legend.key.width = unit(2.2, "cm"),
    legend.spacing.x = unit(0, "cm"),
    plot.margin = margin(t = 80, r = 20, b = 40, l = 20),
    plot.caption = element_text(hjust = 0.5)
  ) +
  guides(fill = guide_legend(ncol = 5, byrow = TRUE)) 

# Title and subtitle annotations
title <- ggdraw() +
  draw_label("INCOME AND EXPENDITURE OF 150 NEGRO FAMILIES IN ATLANTA,GA,USA", 
             fontface = 'bold', x = 0.5, y = 0.65, hjust = 0.5, size = 12) +
  draw_label("Annual Expenditure for    Rent    Food    Clothes    Direct Taxes    Others", 
             x = 0.5, y = 0.45, hjust = 0.5, size = 10)

# Combine title, subtitle, background, and main plot
final_plot <- ggdraw() +
  draw_grob(rasterGrob(bg_img, width = unit(1, "npc"), height = unit(1, "npc"))) +
  draw_plot(title, x = 0, y = 0.85, width = 1, height = 0.15) + # Adjust title positioning
  draw_plot(p)

# Display the final plot
print(final_plot)
```
Interpretion:
This R code creates a historical-style chart that shows how different income classes in Atlanta spent their money on categories like rent, food, clothes, taxes, and other expenses. It processes the data to calculate how much of each class’s total income went to each category, and then displays this using a colorful stacked bar chart. The chart uses a parchment paper background and adds a bold title and subtitle to match the style of W.E.B. Du Bois’s original 1900 design. Each bar represents a class, and percentage labels appear for spending categories that take up more than 1.5% of the budget, making it easy to compare spending habits across income levels.


```

## 2 - COVID survey - interpret

```{r}
#| label: label-me-2

```
Interpret:
In analyzing the survey results on attitudes toward COVID-19 vaccination among medical and nursing students across the U.S., the visualization reveals a range of insights across various demographic groups. Here are three notable patterns, some of which align with intuition and some that deviate.

Age Differences in Perceived Safety and Vaccine Recommendations: In both age categories, the majority of respondents agree with the statement “By my best knowledge I do not have any doubt the vaccine is safe” and would encourage their family, friends and the community at large to take the vaccine. This can be further seen by the decreased Likert values (tending towards 1) and narrow bars of error for these groups in these categories. The youngest respondents (26 years old) exhibit the lowest mean score, while 26-30 and >30 years old respondents indicate somewhat higher mean values and larger error bars, indicating higher opinion variance. This could be interpreted to mean that for some reason, the younger students may for example, due to less likely hood of getting seriously ill with the virus or due to the fact that they are more exposed to information about the vaccine during their studies, may feel more comfortable supporting the vaccine. That brightness can be suggested by the discovery that younger individuals, particularly in medical fields may be more inclined to recommend vaccines.

Gender Variations in Confidence and Concern Over Safety: Some of the variations in attitudes to COVID-19 vaccines are quite subtle but are markedly different between male and female participants. The results reveal fairly high levels of confidence in both the safety of vaccines and the scientific review process among both males and females, as evidenced by mean Likert scale scores of approximately 2. However, there is a slightly higher concern among respondents belonging to the Non Binary / Third gender Category mean scored 2.89 in vaccine safety and 2.9 in the side effect of the vaccine with closer to 3. This is similar to studies done on gender minorities where people of the minority gender are more likely to have a liberal attitude towards health interventions because of the bias that has been accorded to them in the past or unfavourable convictions about medical research. As I expected, gender identify could have an impact on perceptions of new medical technologies but the degree of this effect is barely discernable in this case.

Prior COVID-19 and Flu Vaccination as Predictors of Vaccine Confidence: Exploration of means indicates that prior coverage for COVID-19 vaccine has some students more confident with safety and scientific rigor, currently at mean 1-2. In contrast, students who had not taken the COVID vaccine had intermediate or even negative attitudes, with scores equal to 3. Likewise, those who have taken a flu vaccine this year are relatively more likely to encourage others to take the COVID-19 vaccines. This finding is aligned with the theory that previous uptake behaviour is a great predictor of favourable attitudes towards vaccines because previous experience with vaccinations must mitigate hesitance and build confidence in vaccination.

Altogether, these cases give a picture of how demographic features interconnect in the course of attitude formation toward vaccination. This help to identify the expectant and unexpectant trends in acceptance and with the help of specific segment marketing approaches that may help improve the acceptance level of the vaccine among those groups who may have specific reasons not to accept the vaccine.

## 3 - COVID survey - reconstruct
```{r}
# Load required libraries
library(tidyverse)

# Load and clean the COVID survey data
covid_survey <- read_csv("data/covid-survey.csv", skip = 1) %>%
  filter(!if_all(-response_id, is.na)) %>%
  mutate(
    exp_already_vax = recode(as.character(exp_already_vax), "0" = "No", "1" = "Yes"),
    exp_flu_vax = recode(as.character(exp_flu_vax), "0" = "No", "1" = "Yes"),
    exp_profession = recode(as.character(exp_profession), "0" = "Medical", "1" = "Nursing"),
    exp_gender = recode(as.character(exp_gender), 
                        "0" = "Male", "1" = "Female", 
                        "3" = "Non-binary third gender", 
                        "4" = "Prefer not to say"),
    exp_race = recode(as.character(exp_race),
                      "1" = "American Indian / Alaskan Native", 
                      "2" = "Asian",
                      "3" = "Black / African American", 
                      "4" = "Native Hawaiian / Other Pacific Islander", 
                      "5" = "White"),
    exp_ethnicity = recode(as.character(exp_ethnicity),
                           "1" = "Hispanic / Latino", 
                           "2" = "Non-Hispanic/Non-Latino"),
    exp_age_bin = recode(as.character(exp_age_bin),
                         "0" = "<20", "20" = "21-25", 
                         "25" = "26-30", "30" = ">30")
  )

# Transform to long format
covid_survey_longer <- covid_survey %>%
  pivot_longer(cols = starts_with("exp_"), 
               names_to = "explanatory", 
               values_to = "explanatory_value") %>%
  filter(!is.na(explanatory_value)) %>%
  pivot_longer(cols = starts_with("resp_"), 
               names_to = "response", 
               values_to = "response_value")

# Summary statistics by group (10th and 90th percentiles)
covid_survey_summary_stats_by_group <- covid_survey_longer %>%
  group_by(explanatory, explanatory_value, response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.10, na.rm = TRUE),
    high = quantile(response_value, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

# Summary stats for the full dataset ("All" group)
covid_survey_summary_stats_all <- covid_survey_longer %>%
  group_by(response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.10, na.rm = TRUE),
    high = quantile(response_value, 0.90, na.rm = TRUE),
    explanatory = "All",
    explanatory_value = factor("")
  )

# Combine both
covid_survey_summary_stats <- bind_rows(
  covid_survey_summary_stats_all,
  covid_survey_summary_stats_by_group
)

# Define readable response labels
response_labels <- c(
  resp_safety = "Based on my understanding, I believe the vaccine is safe",
  resp_confidence_science = "I am confident in the scientific vetting process for the new COVID vaccines",
  resp_concern_safety = "I am concerned about the safety and side effects of the vaccine",
  resp_feel_safe_at_work = "Getting the vaccine will make me feel safer at work",
  resp_will_recommend = "I will recommend the vaccine to family, friends, and community members",
  resp_trust_info = "I trust the information that I have received about the vaccines"
)

# Plot
ggplot(covid_survey_summary_stats, aes(x = explanatory_value, y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = low, ymax = high), width = 0.2) +
  facet_grid(response ~ explanatory, scales = "free_x",
             labeller = labeller(
               response = as_labeller(response_labels),
               explanatory = label_wrap_gen(15)
             )) +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "gray90", color = NA),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "Mean Response Value")


```


Interpretion:
This plot visualizes how different demographic and experiential groups responded to COVID-19 vaccine-related survey questions, using mean Likert-scale ratings and error bars spanning the 10th to 90th percentiles. Each point represents the average response for a specific subgroup (e.g., Female, Nursing, or Age 21–25) to a given question, allowing us to compare attitudes across groups. For instance, we can observe whether trust in vaccine information or concern about side effects varies by age, gender, or professional background. The inclusion of overall averages (“All”) provides a baseline for comparison. Overall, the plot highlights both the central tendencies and variability in vaccine perceptions across key explanatory variables.


## 4 - COVID survey - re-reconstruct



```{r}
```{r}
library(tidyverse)
# Recalculate summary statistics with 25th and 75th percentiles
covid_survey_summary_iqr <- covid_survey_longer %>%
  group_by(explanatory, explanatory_value, response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Add "All" category (unconditioned summary stats)
covid_survey_summary_all_iqr <- covid_survey_longer %>%
  group_by(response) %>%
  summarise(
    mean = mean(response_value, na.rm = TRUE),
    low = quantile(response_value, 0.25, na.rm = TRUE),
    high = quantile(response_value, 0.75, na.rm = TRUE),
    explanatory = "All",
    explanatory_value = factor(""),
    .groups = "drop"
  )

# Combine both
covid_survey_summary_stats_iqr <- bind_rows(
  covid_survey_summary_all_iqr,
  covid_survey_summary_iqr
)

# Define readable labels again
response_labels <- c(
  resp_safety = "Based on my understanding, I believe the vaccine is safe",
  resp_feel_safe_at_work = "Getting the vaccine will make me feel safer at work",
  resp_concern_safety = "I am concerned about the safety and side effects of the vaccine",
  resp_confidence_science = "I am confident in the scientific vetting process for the new COVID vaccines",
  resp_trust_info = "I trust the information that I have received about the vaccines",
  resp_will_recommend = "I will recommend the vaccine to family, friends, and community members"
)

explanatory_labels <- c(
  exp_age_bin = "Age",
  exp_gender = "Gender",
  exp_race = "Race",
  exp_ethnicity = "Ethnicity",
  exp_profession = "Profession",
  exp_already_vax = "Had COVID vaccine",
  exp_flu_vax = "Had flu vaccine this year",
  All = "All"
)

# Plot using IQR (25th–75th)
ggplot(covid_survey_summary_stats_iqr, aes(x = mean, y = fct_rev(explanatory_value))) +
  geom_point() +
  geom_errorbarh(aes(xmin = low, xmax = high), height = 0.2) +
  facet_grid(
    fct_rev(explanatory) ~ response,
    scales = "free_y",
    space = "free_y",
    labeller = labeller(
      response = response_labels,
      explanatory = explanatory_labels
    )
  ) +
  scale_x_continuous(breaks = 1:5) +
  labs(
    x = "Mean Likert score",
    y = NULL,
    caption = "Error bars range from 25th to 75th percentile"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.spacing = unit(1, "lines")
  )
```




```

```
Interpretion: 
 This revised plot gives a more robust view of variability by focusing on the interquartile range (middle 50%), rather than the broader 80% used in Question 2. As a result: The error bars are shorter, reducing the impression of high variability in some groups. Outliers are de-emphasized, offering a more stable picture of how responses vary. Some conclusions from Q2 about group differences may appear more stable or less extreme. Conclusion: Using the IQR makes the chart more interpretable and aligned with best practices for Likert-type data. It does not radically change the key findings but strengthens confidence in central tendencies.



## 5 - COVID survey - another view
```{r}
# Load required libraries
library(tidyverse)
library(forcats)
library(ggplot2)
# Prepare Likert-style sentiment data
likert_data <- covid_survey %>%
  select(starts_with("resp_")) %>%
  pivot_longer(everything(), names_to = "question", values_to = "response") %>%
  mutate(
    question = fct_recode(question,
      "Believe safe" = "resp_safety",
      "Trust science" = "resp_confidence_science",
      "Concerned" = "resp_concern_safety",
      "Feel safe" = "resp_feel_safe_at_work",
      "Would recommend" = "resp_will_recommend",
      "Trust info" = "resp_trust_info"
    ),
    sentiment = case_when(
      response < 3 ~ "Positive",
      response > 3 ~ "Negative",
      TRUE ~ "Neutral"
    )
  )

# Plot diverging bar chart
ggplot(likert_data, aes(x = question, fill = sentiment)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Positive" = "#1f78b4", "Negative" = "#e31a1c", "Neutral" = "gray70")) +
  coord_flip() +
  labs(
    title = "Diverging Bar Chart of Vaccine Attitudes",
    x = NULL,
    y = "Count of Responses",
    fill = "Sentiment"
  ) +
  theme_minimal()

ggplot(likert_data, aes(x = question, fill = factor(response))) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "RdYlBu",
    labels = c("1: Strongly Agree", "2: Agree", "3: Neutral",
               "4: Disagree", "5: Strongly Disagree")) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(
    title = "100% Stacked Bar Chart",
    x = NULL,
    y = "Percentage of Responses",
    fill = "Response Scale"
  ) +
  theme_minimal()

```
5C.
The visualization shows the distribution of responses for each question, making it simple to visually compare percentage distributions. The bars extend in opposite directions, highlighting response polarity and clearly representing the ‘agree-disagree’ scale for each question.

100% Stacked Bar: This chart displays the proportional breakdown of each response option by question, offering a normalized view of response percentages. It effectively shows the composition of response categories within each question.

Comparability: The diverging chart emphasizes response ranges, making it easier to compare overall agreement or disagreement between questions. On the other hand, the 100% stacked chart normalizes responses, focusing on the proportion of each response category. While the stacked format shows response distribution, it requires more effort to gauge agreement or disagreement between questions compared to the diverging chart.
